---
title: "Differential Expression"
author: "Axel Zagal"
date: "24/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# project list

ERP109920

SRP126648 dopaminergic

ERP111522 HDAC4

ERP017126 pancreas dm2

SRP158367 ipsc

# Differential Expression analysis of ipsc-derived neurons

## We download the selected data set

```{r include=FALSE}
library(recount3)
```


```{r message=FALSE}
human_projects <- available_projects()
proj_info <- subset(
    human_projects,
    project == "SRP158367" & project_type == "data_sources"
)

```
## This creates the rse object using the data set

```{r message=FALSE}
rse_ipsc <- create_rse(proj_info)
```

## We change to a counts per read format

```{r}
assay(rse_ipsc, "counts") <- compute_read_counts(rse_ipsc)

```

## Facilitate the access to data attributes

```{r}
rse_gene_ipsc <- expand_sra_attributes(rse_ipsc)
colData(rse_ipsc)[
    ,
    grepl("^sra_attribute", colnames(colData(rse_ipsc)))
]

```

## Change the data to a factor or numeric data type as needed

```{r}
rse_gene_ipsc$sra_attribute.batch <- as.numeric(rse_gene_ipsc$sra_attribute.batch)
rse_gene_ipsc$sra_attribute.cell_line <- factor(rse_gene_ipsc$sra_attribute.cell_line)
rse_gene_ipsc$sra_attribute.cell_type <- factor(rse_gene_ipsc$sra_attribute.cell_type)
rse_gene_ipsc$sra_attribute.site <- factor(rse_gene_ipsc$sra_attribute.site)
rse_gene_ipsc$sra_attribute.source_name <- factor(rse_gene_ipsc$sra_attribute.source_name)
rse_gene_ipsc$sra_attribute.time_point <- factor(rse_gene_ipsc$sra_attribute.time_point)
rse_gene_ipsc$sra_attribute.well <- factor(rse_gene_ipsc$sra_attribute.well)

```

## Now we use edgeR to normalize the data

```{r}
library("edgeR") 
dge <- DGEList(
    counts = assay(rse_gene_ipsc, "counts"),
    genes = rowData(rse_gene_ipsc)
)
dge <- calcNormFactors(dge)
```

## Get cell lines names

```{r}
summary(as.data.frame(colData(rse_gene_ipsc)[
    ,
    grepl("^sra_attribute.[cell_line]", colnames(colData(rse_gene_ipsc)))
]))
```
## Create a factor of both cell lines

```{r}
rse_gene_ipsc$line <- factor(ifelse(rse_gene_ipsc$sra_attribute.cell_line == "SB808", "A", "B"))
table(rse_gene_ipsc$line)
```

## Get the assigned gene prop

```{r}
rse_gene_ipsc$assigned_gene_prop <- rse_gene_ipsc$recount_qc.gene_fc_count_all.assigned / rse_gene_ipsc$recount_qc.gene_fc_count_all.total

summary(rse_gene_ipsc$assigned_gene_prop)
```

## Compare the assigned gene prop on each cell line

```{r}
library("ggplot2")
ggplot(as.data.frame(colData(rse_gene_ipsc)), aes(y = assigned_gene_prop, x = line)) +
    geom_boxplot() +
    theme_bw(base_size = 20) +
    ylab("Assigned Gene Prop") +
    xlab("Cell Line")
```
## We create a model matrix

```{r}
mod <- model.matrix(~ line +  + assigned_gene_prop,
    data = colData(rse_gene_ipsc)
)

```
## This creates a voom plot using the dge object and the model matrix

```{r}
library("limma")
vGene <- voom(dge, mod, plot = TRUE)
```

## Then we use the eBayes function for differential expression

```{r}
eb_results <- eBayes(lmFit(vGene))

de_results <- topTable(
    eb_results,
    coef = 2,
    number = nrow(rse_gene_ipsc),
    sort.by = "none"
)
dim(de_results)
```

## Number of genes expressed in each group with FDR < 0.05%

```{r}
table(de_results$adj.P.Val < 0.05)
```

## We can visualize the results using plotMA and a volcano plot

```{r}
plotMA(eb_results, coef = 2)
```

```{r}
volcanoplot(eb_results, coef = 2, highlight = 3, names = de_results$gene_name)
```

## Finally, we display the three highlighted genes

```{r}
de_results[de_results$gene_name %in% c("TMSB4XP8", "TMSB4X", "ANKRD20A"), ]
```
